---
title: "Spatial Machine Learning for Air Pollution Modeling in Poland"
author: "Ondrej Marvan & Adam Jaworski"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
    fig_width: 10
    fig_height: 7
  pdf_document:
    toc: true
subtitle: "Predicting Ground-Level NO2 Using Sentinel-5P TROPOMI & Spatial Features"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, fig.align="center", cache=TRUE)
source("00_setup.R")
library(sf); library(terra); library(tidyverse); library(spdep)
library(ranger); library(gstat); library(automap)
library(dbscan); library(cluster); library(viridis); library(corrplot); library(knitr)
```

# 1. Introduction

Air pollution, particularly nitrogen dioxide (NO2), is a major public health concern in Poland. This study combines **Sentinel-5P TROPOMI** satellite observations with **GIOS** ground station data to build spatial ML models predicting NO2 concentrations at 5 km resolution across Poland.

# 2. Data

```{r load-data}
grid <- safe_read_sf(file.path(CONFIG$dir_processed, "analysis_grid.gpkg"))
tropomi <- safe_read_csv(file.path(CONFIG$dir_processed, "tropomi_grid.csv"))
stations <- safe_read_sf(file.path(CONFIG$dir_processed, "stations.gpkg"))
st_annual <- safe_read_csv(file.path(CONFIG$dir_processed, "no2_station_annual.csv"))
```

| Source | Content | Resolution |
|--------|---------|-----------|
| Sentinel-5P TROPOMI | NO2 tropospheric columns | ~5.5 km, fortnightly |
| GIOS | Ground NO2, SO2, PM10 | ~250 stations, hourly |
| GUS BDL | Population | Gmina level |
| OpenStreetMap | Roads, railways, industry | Vector |

# 3. EDA

```{r no2-map, fig.cap="Annual mean tropospheric NO2."}
if (!is.null(grid) && "tropomi_no2_mean" %in% names(grid)) {
  ggplot(grid |> filter(!is.na(tropomi_no2_mean))) +
    geom_sf(aes(fill = tropomi_no2_mean), color = NA) +
    scale_fill_viridis(option = "inferno", name = expression(NO[2])) +
    labs(title = "Annual Mean Tropospheric NO2 - Poland 2024") + theme_minimal()
}
```

```{r timeseries, fig.cap="Temporal NO2 pattern."}
if (!is.null(tropomi)) {
  tropomi |> group_by(obs_date) |>
    summarise(mean_no2 = mean(tropomi_no2, na.rm=TRUE), .groups="drop") |>
    ggplot(aes(obs_date, mean_no2)) + geom_line(color="steelblue", linewidth=1) +
    labs(x=NULL, y="NO2 density") + theme_minimal()
}
```

```{r correlations, fig.cap="Feature correlations."}
if (!is.null(grid)) {
  nc <- grid |> st_drop_geometry() |>
    select(where(is.numeric), -cell_id, -x_coord, -y_coord) |> drop_na()
  if (ncol(nc) >= 3 && nrow(nc) > 30)
    corrplot(cor(nc), method="color", type="lower", tl.cex=0.7, addCoef.col="black", number.cex=0.6)
}
```

```{r moran}
if (!is.null(grid) && "tropomi_no2_mean" %in% names(grid)) {
  cl <- grid |> filter(!is.na(tropomi_no2_mean))
  co <- st_coordinates(st_centroid(cl))
  lw <- nb2listw(knn2nb(knearneigh(co, k=8)), style="W")
  mi <- moran.test(cl$tropomi_no2_mean, lw)
  cat(sprintf("Moran's I = %.4f (p = %s)\n", mi$estimate[1], format.pval(mi$p.value)))
}
```

# 4. Supervised ML

```{r rf-importance, fig.cap="Random Forest variable importance."}
f <- file.path(CONFIG$dir_figures, "rf_importance.png")
if (file.exists(f)) knitr::include_graphics(f)
```

```{r model-comparison}
f <- file.path(CONFIG$dir_output, "model_comparison.csv")
if (file.exists(f)) read_csv(f, show_col_types=FALSE) |> kable(digits=4)
```

```{r predicted-map, fig.cap="Predicted ground-level NO2."}
f <- file.path(CONFIG$dir_figures, "predicted_no2.png")
if (file.exists(f)) knitr::include_graphics(f)
```

# 5. Unsupervised ML

```{r hotspots, fig.cap="DBSCAN hotspot clusters."}
f <- file.path(CONFIG$dir_figures, "hotspots.png")
if (file.exists(f)) knitr::include_graphics(f)
```

```{r lisa, fig.cap="LISA cluster map."}
f <- file.path(CONFIG$dir_figures, "lisa.png")
if (file.exists(f)) knitr::include_graphics(f)
```

```{r pam, fig.cap="PAM spatial typology."}
f <- file.path(CONFIG$dir_figures, "pam_clusters.png")
if (file.exists(f)) knitr::include_graphics(f)
```

# 6. Kriging

```{r variogram, fig.cap="Semivariogram."}
f <- file.path(CONFIG$dir_figures, "variogram.png")
if (file.exists(f)) knitr::include_graphics(f)
```

```{r kriging-map, fig.cap="Ordinary Kriging surface."}
f <- file.path(CONFIG$dir_figures, "kriging.png")
if (file.exists(f)) knitr::include_graphics(f)
```

```{r kriging-cv}
f <- file.path(CONFIG$dir_output, "kriging_comparison.csv")
if (file.exists(f)) read_csv(f, show_col_types=FALSE) |> kable(digits=4)
```

# 7. Discussion

*Key findings to be added after analysis.*

# 8. References

- Copernicus Sentinel-5P, TROPOMI L2 NO2 products
- GIOS, Panstwowy Monitoring Srodowiska
- GUS, Bank Danych Lokalnych
- OpenStreetMap contributors

---
*Spatial ML in R, WNE UW 2026 | prof. K. Kopczewska*

```{r session}
sessionInfo()
```
